---
title: "Statistics and ML"
subtitle: "MSSP Practicum Discussion"
author: "Hsueh-Pin Liu"
date: "2023-01-30"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
```

## Instructions

**Fork** the [`carvalho/stats-ml-practicum`](https://github.com/carvalho/stats-ml-practicum) repository at GitHub, and
**create a new branch with your BU login** to store your changes to the document.
Start by changing the `author`in the YAML header of the document to state **your name**.

Below we run some analyses and ask questions about them. As you run the code and
interpret the results within your group, write your answers to the questions following the analyses, but:

> You should submit your work as a **pull request** to the original repository!


## Introduction

In this project we study **tree canopy cover** as it varies with the
**relative distance** to a tree line boundary in urban forests. The dataset in
`stats-ml-canopy.RData` has three variables: `location` for the urban forest
where the canopy cover was observed, `distance` for the relative distance &mdash;
zero is inside the forest and one is outside (city) &mdash; and `cover` for the
canopy cover.

```{r}
load("stats-ml-canopy.RData")
(canopy <- as_tibble(canopy))

idx <- order(canopy$distance) # for plots below
ggplot(canopy, aes(distance, cover)) + geom_point(color = "gray")
```

As can be seen, there is a clear pattern here: the canopy cover starts high,
closer to 100% when inside the forest, but as the tree line recedes into the
city, the canopy cover approaches zero.

We are interested in two main tasks:

- **Understanding** this relationship more explicitly;
- **Predicting** the canopy cover at the assumed tree line boundary when
`distance` is 0.5.

To this end, we explore four approaches below.

## Statistics 1: Linear Fit

```{r stats1}
m <- glm(cover ~ distance, data = canopy, family = quasibinomial)
ggplot(canopy, aes(distance, cover)) + geom_point(col = "gray") +
  geom_line(aes(distance[idx], fitted(m)[idx]))
predict(m, data.frame(distance = 0.5), se = TRUE, type = "response")
plot(m)
```

Questions and tasks:

- Comment on the fit, plot residuals and comment on them.
- Comment on the prediction; does it seem reasonable?

Though the fit is not correct, it still gives a brief figure of the model.
The fit is close to 0.5, so it seems reasonable, but remain to be proved.



## ML 1: LOESS

```{r ml1}
m <- loess(cover ~ distance, data = canopy)
ggplot(canopy, aes(distance, cover)) + geom_point(col = "gray") +
  geom_line(aes(distance[idx], fitted(m)[idx]))
predict(m, data.frame(distance = 0.5), se = TRUE)
```

Questions and tasks:

- Check the definition of the `loess` function; how does it differ from the previous approach?
- Comment on the fit; does it seem reasonable?
- Comment on the prediction, including the SE.

It divides the data into several small parts, repeating this process and finally get the result.
The curve looks like the previous one, but for both tails, it seems a bit unreasonable because when you are in the forest, the cover should be one and when you are away from the forest, it should be 0.
The fit is still at about 50%, and the SE is a bit smaller than the previous one, so the model has a better result.


## ML 2: Random Forest

```{r ml2,message=FALSE}
library(randomForest)
m <- randomForest(cover ~ distance, data = canopy)
ggplot(canopy, aes(distance, cover)) + geom_point(col = "gray") +
  geom_line(aes(distance[idx], predict(m)[idx]))
predict(m, data.frame(distance = 0.5), se = TRUE)
```

Questions and tasks:

- Check what `randomForest` does; what is **keyword** here?
- Comment on the fit; how does it differ from the previous fits?
- Comment on the prediction; how would you obtain a measure of uncertainty?

It randomly selects data and make groups for them,so normally it gets a higher accuracy and can deal with a large dataset.
Obviously the line is not a curve anymore, but a combination of many lines.
The prediction is quite different from the previous two. First I should identify the sources of uncertainty in the measurement,then estimate the size, finally they are combined to give an overall figure.

## Statistics 2: Cubic Fit

```{r stats2}
m <- glm(cover ~ poly(distance, 3), data = canopy, family = quasibinomial)
ggplot(canopy, aes(distance, cover)) + geom_point(col = "gray") +
  geom_line(aes(distance[idx], fitted(m)[idx]))
predict(m, data.frame(distance = 0.5), se = TRUE, type = "response")
plot(m)
```

Questions and tasks:

- Comment on the fit and compare it to the first model; plot and check residuals.
- Comment on the prediction and compare it to previous results.
- How would you know that a cubic fit is good enough?

It seems similar as the first one, but it's actually more fit.
The prediction is back to about 50%.
When the residuals are relatively small.



## Discussion

Let's try to connect all lessons learned from your work and the discussions.
Elaborate more on the following questions:

- How would you know that the predictions are *reliable*?
- How would you test that the cover is exactly 50% at the boundary (`distance` = 0.5)? Which approaches would make the test easier to perform?
- How would you incorporate `location` in your analyses? How would you know that
it is meaningful to use it?

After checking the models, plot and residuals, if it shows small residuals and reasonable results, it's reliable.
By using distance=0.5 in the model and find out results.
Location actually help us group up all the data, we can use methods of resampling to better check whether the model fits.
