---
title: "Statistics and ML"
subtitle: "MSSP Practicum Discussion"
author: "MSSP Practicum II"
date: "2023-01-24"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
```

## Instructions

**Fork** the [`carvalho/stats-ml-practicum`](https://github.com/carvalho/stats-ml-practicum) repository at GitHub, and
**create a new branch with your BU login** to store your changes to the document.
Start by changing the `author`in the YAML header of the document to state **your name**.

Below we run some analyses and ask questions about them. As you run the code and
interpret the results within your group, write your answers to the questions following the analyses, but:

> You should submit your work as a **pull request** to the original repository!


## Introduction

In this project we study **tree canopy cover** as it varies with the
**relative distance** to a tree line boundary in urban forests. The dataset in
`stats-ml-canopy.RData` has three variables: `location` for the urban forest
where the canopy cover was observed, `distance` for the relative distance &mdash;
zero is inside the forest and one is outside (city) &mdash; and `cover` for the
canopy cover.

```{r}
load("stats-ml-canopy.RData")
(canopy <- as_tibble(canopy))

idx <- order(canopy$distance) # for plots below
ggplot(canopy, aes(distance, cover)) + geom_point(color = "gray")
```

As can be seen, there is a clear pattern here: the canopy cover starts high,
closer to 100% when inside the forest, but as the tree line recedes into the
city, the canopy cover approaches zero.

We are interested in two main tasks:

- **Understanding** this relationship more explicitly;
- **Predicting** the canopy cover at the assumed tree line boundary when
`distance` is 0.5.

To this end, we explore four approaches below.

## Statistics 1: Linear Fit

```{r stats1}
m <- glm(cover ~ distance, data = canopy, family = quasibinomial)
ggplot(canopy, aes(distance, cover)) + geom_point(col = "gray") +
  geom_line(aes(distance[idx], fitted(m)[idx]))
predict(m, data.frame(distance = 0.5), se = TRUE, type = "response")
```
```{r}
plot(m)
```


Questions and tasks:

- Comment on the fit, plot residuals and comment on them.

Since the residuals are not uniformly and randomly distributed, but there are some patterns, I think the model does not fit very well.

- Comment on the prediction; does it seem reasonable?


## ML 1: LOESS

```{r ml1}
m <- loess(cover ~ distance, data = canopy)
ggplot(canopy, aes(distance, cover)) + geom_point(col = "gray") +
  geom_line(aes(distance[idx], fitted(m)[idx]))
predict(m, data.frame(distance = 0.5), se = TRUE)
```

Questions and tasks:

- Check the definition of the `loess` function; how does it differ from the previous approach?

loess function is used for fitting a polynomial surface determined by one or more numerical predictors, using local fitting.

In local polynomial regression fitting, when the value of a response variable is to be estimated, a subset of data is first taken from the vicinity of its predictor variable, and then linear or quadratic regression is performed on this subset, and the weighted least squares method is used in the regression.

- Comment on the fit; does it seem reasonable?

```{r}
plot(m$resid~m$fit)
```

- Comment on the prediction, including the SE.


## ML 2: Random Forest

```{r ml2,message=FALSE}
library(randomForest)
m <- randomForest(cover ~ distance, data = canopy)
ggplot(canopy, aes(distance, cover)) + geom_point(col = "gray") +
  geom_line(aes(distance[idx], predict(m)[idx]))
predict(m, data.frame(distance = 0.5), se = TRUE)
```




Questions and tasks:

- Check what `randomForest` does; what is **keyword** here?

Implements Breiman's random forest algorithm (based on Breiman and Cutler's original Fortran code) for classification and regression. It can also be used in unsupervised mode for assessing proximities among data points.

- Comment on the fit; how does it differ from the previous fits?

```{r}
plot(m)
```
The model is predicted with high accuracy, with no need for further tuning. 

Its avoids overfitting.

- Comment on the prediction; how would you obtain a measure of uncertainty?




## Statistics 2: Cubic Fit

```{r stats2}
m <- glm(cover ~ poly(distance, 3), data = canopy, family = quasibinomial)
ggplot(canopy, aes(distance, cover)) + geom_point(col = "gray") +
  geom_line(aes(distance[idx], fitted(m)[idx]))
predict(m, data.frame(distance = 0.5), se = TRUE, type = "response")
```

Questions and tasks:

- Comment on the fit and compare it to the first model; plot and check residuals.

```{r}
plot(m)
```

- Comment on the prediction and compare it to previous results.
- How would you know that a cubic fit is good enough?

Assuming the model you fit to the data is correct, the residuals approximate the random errors. Therefore, if the residuals appear to behave randomly, it suggests that the model fits the data well. However, if the residuals display a systematic pattern, it is a clear sign that the model fits the data poorly.

## Discussion

Let's try to connect all lessons learned from your work and the discussions.
Elaborate more on the following questions:

- How would you know that the predictions are *reliable*?
- How would you test that the cover is exactly 50% at the boundary (`distance` = 0.5)? Which approaches would make the test easier to perform?
- How would you incorporate `location` in your analyses? How would you know that
it is meaningful to use it?
